---
title: "PS5 Assignment"
author: "Nawal Dhabar"
date: "2023-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

## Gapminder data
We use gapminder dataset, downloaded from https://www.gapminder.org/data/, however, the data structure there is quire complex, please use the dataset provided on canvas (in files/data).
The variables are:
_name_: country name
_iso3_: 3-letter country code
_iso2_: 2-letter country code
_region_: broad geographic region
_sub-region_: more precise region
_intermediate-region_
_time_: year
_totalPopulation_: total population
_GDP_PC_: GDP per capita (constant 2010 US$)
_accessElectricity_: Access to electricity (% of population)
_agriculturalLand_: Agricultural land (sq. km)
_agricultureTractors_: Agricultural machinery, tractors (count)
_cerealProduction_: Cereal production (metric tons)
_feritilizerHa_: Fertilizer consumption (kilograms per hectare of arable land)
_fertilityRate_: total fertility rate (births per woman)
_lifeExpectancy_: Life expectancy at birth, total (years)
_childMortality_: Mortality rate, under-5 (per 1,000 live births)
_youthFemaleLiteracy_: Literacy rate, youth female (% of females ages 15-24)
_youthMaleLiteracy_: Literacy rate, youth male (% of males ages 15-24)
_adultLiteracy_: Literacy rate, adult total (% of people ages 15 and above)
_co2_: CO2 emissions (kt)
_greenhouseGases_: Total greenhouse gas emissions (kt of CO2 equivalent)
_co2_PC_: CO2 emissions (metric tons per capita)
_*'pm2.5_35'*_: PM2.5 pollution, population exposed to levels exceeding WHO Interim Target-1 value 36ug/m3 (
_battleDeaths_: Battle-related deaths (number of people)

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(stringr)
```

## 1 Load and check data (5pt)
You first task is to do a very simple data check:
1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/
r-markdown.htmlr-markdown-rstudio-creating.

2. (2pt) Load data. How many rows/columns do we have?
```{r, echo=FALSE, message=FALSE}
gapminder <- read_delim('/Users/nawald/Desktop/gapminder.csv.bz2')
```

```{r, echo=FALSE}
row <- nrow(gapminder)
column <- ncol(gapminder)
```
We have `r row` rows and `r column` columns in the data


3. (2pt) Print a small sample of data. Does it look OK?
```{r, echo=FALSE}
gapminder %>% 
sample_n(3) %>% 
  knitr::kable()
```
The data looks OK.

## 2 Descriptive statistics (15pt)
1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r}
countries_iso3 <- gapminder %>% 
  select(iso3) %>% 
  filter(!is.na(iso3)) %>% 
  n_distinct()

countries_iso2 <- gapminder %>% 
  select(iso2) %>% 
  filter(!is.na(iso2)) %>% 
  n_distinct() 

countries_name <- gapminder %>% 
  select(name) %>% 
  filter(!is.na(name)) %>% 
  n_distinct() 
```
There are `r countries_iso3` countries according to the iso3 variable, `r countries_iso2` countries according to my calculations of the iso2 variable and `r countries_name` countries according to the name variable



2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?
  (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?

```{r, echo=FALSE} 
find_out <- gapminder %>% 
  group_by(iso2) %>% 
  summarize(total = n_distinct(name)) %>% 
  arrange(-total)

which <- gapminder %>% 
  select(iso2, name) %>% 
  filter(is.na(iso2)) %>% 
  distinct()

```
Yes, there is an iso-2 code that corresponds to more than one name, this code is called `r which [1, 1]`. It corresponds to two country names. One of these country names is   `r which[2, 2]` and the other is called NA, most likely because it does not have country name stored.


  (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.

``` {r}
thing <- gapminder %>% 
  group_by(name) %>% 
  summarize(total = n_distinct(iso3)) %>% 
  arrange(-total)

another <- gapminder %>% 
  select(iso3, name) %>% 
  filter(is.na(name)) %>% 
  distinct()

```
Yes, there are country names that have more than one iso3-code. The country name is stored as 'NA' in the data. It has four iso3 codes assigned to it, these codes are `r another[1,1]`, `r another[2,1]`, `r another[3, 1]`, and `r another[4,1]`. I believe these codes were not assigned names, which is why their corresponding values in the name column is 'NA'

3. (2pt) What is the minimum and maximum year in these data?

```{r}
numbers <- gapminder %>% 
  filter(!is.na(time)) 
  
mini <- min(numbers$time)
maxi <- max(numbers$time)

```
The maximum year in the data is `r maxi ` and the minimum year is `r mini`


## 3 CO2 emissions (30pt)
Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

```{r, echo=FALSE}
 missing_co2 <- gapminder %>% 
  filter(is.na(co2)) %>% 
  filter(!is.na(time)) %>% 
  group_by(time) %>% 
  filter(time %in% (1960:2019)) %>% 
  summarize(
    missing_co2_values = n()
  )  %>% 
   filter(rank(desc(missing_co2_values)) < 4) %>% 
  knitr::kable() 
  
```
These are the missing co2 emissions for each year, using the co2 column to calculate: `r missing_co2`. The years with the most missing data here are 2019, 2018 and 2017. 


``` {r, echo=FALSE}
missing_co2_PC <- gapminder %>% 
  filter(is.na(co2_PC)) %>% 
  filter(!is.na(time)) %>% 
  group_by(time) %>% 
  filter(time %in% (1960:2019)) %>% 
  summarize(
    missing_co2_PC_values = n()
  ) %>% 
  filter(rank(desc(missing_co2_PC_values)) < 4) %>% 
  knitr::kable()
```

These are the missing co2 emissions for each year, using the co2 per capita column to calculate: `r missing_co2_PC`. Once again, the years with the most missing data are 2019, 2018 and 2017.



2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.



```{r, echo=FALSE}
emissions <- gapminder %>% 
      group_by(name) %>% 
      filter(!is.na(co2)) %>% 
      filter(!is.na(time)) %>% 
      filter(name == 'United States of America'| name == 'China' | name == 'India'
             | name == 'Aruba' | name == 'Afghanistan') %>% 
      summarize(
        total_co2 = sum(co2)
      )

ggplot(emissions, aes(name, total_co2)) +
      geom_col(fill = 'mediumpurple1', 
              linewidth=2)

```


I see that Afghanistan and Aruba have a significantly less amount of total co2 emissions than the U.S., China and India, to the point where you can barely see their bars on the chart. I think this is because of the gap in development between these two groups of countries is big, so the less developed ones produce less co2 emissions. They are also less populated than the countries with higher co2 emissions, so that also plays a part. 

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the same countries. What does this figure suggest?

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different size.

Hint: Americas 2016 should be 4.80.
5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).


## 4 GDP per capita (50pt)
Let’s look at GDP per capita (GDP_PC ).

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.

2. (4pt) Make a similar plot, but this time use 2019 data only.

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of different size.
5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!

7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)
  Hint: check out the function rank()!
  Hint2: 17 for 1960.
  
8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? Hint: 0.0904 for 1960.

## Finally tell us how many hours did you spend on this PS.

