---
title: "PS5 Assignment"
author: "Nawal Dhabar"
date: "2023-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(stringr)
```

## 1 Load and check data (5pt)
You first task is to do a very simple data check:
1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/
r-markdown.htmlr-markdown-rstudio-creating.

2. (2pt) Load data. How many rows/columns do we have?
```{r, echo=FALSE, message=FALSE}
gapminder <- read_delim('/Users/nawald/Desktop/gapminder.csv.bz2')
```

```{r, message=FALSE}
row <- nrow(gapminder)
column <- ncol(gapminder)
```
We have `r row` rows and `r column` columns in the data


3. (2pt) Print a small sample of data. Does it look OK?
```{r, message=FALSE}
gapminder %>% 
sample_n(3) %>% 
  knitr::kable()
```
The data looks OK.

## 2 Descriptive statistics (15pt)
1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r, message=FALSE}
countries_iso3 <- gapminder %>% 
  select(iso3) %>% 
  filter(!is.na(iso3)) %>% 
  n_distinct()

countries_iso2 <- gapminder %>% 
  select(iso2) %>% 
  filter(!is.na(iso2)) %>% 
  n_distinct() 

countries_name <- gapminder %>% 
  select(name) %>% 
  filter(!is.na(name)) %>% 
  n_distinct() 
```
There are `r countries_iso3` countries according to the iso3 variable, `r countries_iso2` countries according to my calculations of the iso2 variable and `r countries_name` countries according to the name variable



2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?
  (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?

```{r, message=FALSE} 
find_out <- gapminder %>% 
  group_by(iso2) %>% 
  summarize(total = n_distinct(name)) %>% 
  arrange(-total)

which <- gapminder %>% 
  select(iso2, name) %>% 
  filter(is.na(iso2)) %>% 
  distinct()

```
Yes, there is an iso-2 code that corresponds to more than one name, this code is called `r which [1, 1]`. It corresponds to two country names. One of these country names is   `r which[2, 2]` and the other is called NA, most likely because it does not have country name stored.


  (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.

``` {r, message=FALSE}
thing <- gapminder %>% 
  group_by(name) %>% 
  summarize(total = n_distinct(iso3)) %>% 
  arrange(-total)

another <- gapminder %>% 
  select(iso3, name) %>% 
  filter(is.na(name)) %>% 
  distinct()

```
Yes, there are country names that have more than one iso3-code. The country name is stored as 'NA' in the data. It has four iso3 codes assigned to it, these codes are `r another[1,1]`, `r another[2,1]`, `r another[3, 1]`, and `r another[4,1]`. I believe these codes were not assigned names, which is why their corresponding values in the name column is 'NA'

3. (2pt) What is the minimum and maximum year in these data?

```{r, message=FALSE}
numbers <- gapminder %>% 
  filter(!is.na(time)) 
  
mini <- min(numbers$time)
maxi <- max(numbers$time)

```
The maximum year in the data is `r maxi ` and the minimum year is `r mini`


## 3 CO2 emissions (30pt)
Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

```{r, message=FALSE}
 missing_co2 <- gapminder %>% 
  filter(is.na(co2)) %>% 
  filter(!is.na(time)) %>% 
  group_by(time) %>% 
  filter(time %in% (1960:2019)) %>% 
  summarize(
    missing_co2_values = n()
  )  %>% 
   filter(rank(desc(missing_co2_values)) < 4) %>% 
  knitr::kable() 
  
```
These are the missing co2 emissions for each year, using the co2 column to calculate: `r missing_co2`
The years with the most missing data here are 2019, 2018 and 2017. 


``` {r, message=FALSE}
missing_co2_PC <- gapminder %>% 
  filter(is.na(co2_PC)) %>% 
  filter(!is.na(time)) %>% 
  group_by(time) %>% 
  filter(time %in% (1960:2019)) %>% 
  summarize(
    missing_co2_PC_values = n()
  ) %>% 
  filter(rank(desc(missing_co2_PC_values)) < 4) %>% 
  knitr::kable()
```

These are the missing co2 emissions for each year, using the co2 per capita column to calculate: `r missing_co2_PC`
Once again, the years with the most missing data are 2019, 2018 and 2017.



2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.



```{r, message=FALSE}
emissions <- gapminder %>% 
      group_by(name, time) %>% 
      filter(!is.na(co2)) %>% 
      filter(!is.na(time)) %>% 
      filter(name == 'United States of America'| name == 'China' | name == 'India'
             | name == 'Aruba' | name == 'Afghanistan') %>% 
      summarize(
        total_co2 = sum(co2)
      )
    
    ggplot(emissions, aes(x = time, y = total_co2, color = name)) +
      geom_line() +
      ggtitle('Total co2 emissions over time')

```


I see that Afghanistan and Aruba have a significantly less amount of total co2 emissions than the U.S., China and India, to the point where their lines are low on the chart. I think this is because of the gap in development between these two groups of countries is big, so the less developed ones produce less co2 emissions. They are also less populated than the countries with higher co2 emissions, so that also plays a part. 

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the same countries. What does this figure suggest?


```{r, message=FALSE}
capita <- gapminder %>% 
      group_by(name, time) %>% 
      filter(!is.na(co2_PC)) %>% 
      filter(!is.na(time)) %>% 
      filter(name == 'United States of America'| name == 'China' | name == 'India'
             | name == 'Aruba' | name == 'Afghanistan') %>% 
      
      summarize(
        total_emissions = sum(co2_PC)
      )
    
    ggplot(capita, aes(x = time, y = total_emissions, color = name)) +
      geom_line() +
      ggtitle('Total co2 emissions per capita')

```

This figure is very different from the first one. Here, Aruba which was the one with the lowest total co2 emissions has trended higher than other countries. However, it also crashed down halfway during the period of time the data covers, and is now trending lower than countries like the U.S. Interestingly enough, Afghanistan which was also one of the lowest before is still the lowest now. I think this figure suggests that at one point, Aruba had more co2 per capita than the other countries, despite lower total co2 emissions. However, something happened and now it has decreased significantly while the others continue to grow.


4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different size.

Hint: Americas 2016 should be 4.80.

```{r, message=FALSE} 
average <- gapminder %>% 
      filter(!is.na(co2_PC),!is.na(region)) %>% 
      group_by(region, time) %>% 
      summarize(
        total = mean(co2_PC)
      )

```

I noticed that with a few exceptions, Americas have the highest average co2 emissions per capita, with it's entries in the top ten highest ranging from the years 1970 to 1985 and the values between 10 and 13 metric tons. Another interesting thing to me is that most of the ten highest averages take place between the 1970s and 1980s, despite the range of years for this data maxing out at 2019. It's weird that such high averages are clustered in the same period.



5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.
```{r, message=FALSE}
average <- gapminder %>% 
      filter(!is.na(co2_PC),!is.na(region)) %>% 
      filter(time == '1960'| time == '2016') %>% 
      group_by(region, time) %>% 
      mutate(avg_co2_PC = mean(co2_PC)) 
    
    ggplot(average, aes(region, avg_co2_PC, fill = factor(time))) +
    geom_bar(stat = 'identity', position = 'dodge') +
      ggtitle('Average co2 emissions per capita across continents in 1960 and 2016')

```


6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).

``` {r, message=FALSE}
three <-  gapminder %>% 
  select(name, region, time, co2_PC) %>% 
  filter(!is.na(co2_PC), !is.na(name), !is.na(region), !is.na(time)) %>% 
  filter(time == '2016') %>% 
  group_by(region, name, time) %>% 
  summarize(
    total = sum(co2_PC)
  ) %>% 
    arrange(-total)
    
  three_highest <- three %>% 
    group_by(region) %>% 
    filter(rank(desc(total)) < 4)
  
  three_lowest <- three %>% 
    group_by(region) %>% 
    filter(rank(desc(-total)) < 4)

```

Since 2019 doesn't have enough data, I'll use 2016 because it's the closest year. I found that the three largest co2 emitters in 2016 for the Americas were `r three_highest[2, 2]`, `r three_highest[8, 2]`, and `r three_highest[10, 2]`. The three smallest were `r three_lowest [10, 2]`, `r three_lowest[5, 2]`, and `r three_lowest[4, 2]`. 

The three largest in Asia were `r three_highest[1, 2]`, `r three_highest[3, 2]`, and 
`r three_highest[4,2]`. The three smallest were `r three_lowest[12, 2]`, `r  three_lowest[9,2]`, and `r  three_lowest[8,2]`.

For Europe, the three largest were `r three_highest[6, 2]`, `r three_highest[9, 2]`, and `r three_highest[11, 2]`. The three lowest were `r three_lowest[3, 2]`, 
`r three_lowest[2, 2]`, and `r three_lowest[1, 2]`.

In Oceania, the countries with the largest co2 emissions were `r three_highest[5, 2]`, 
`r three_highest[7, 2]`, and `r three_highest[12, 2]`. The three lowest emitters in this region are `r three_lowest[11, 2]`, `r three_lowest[7, 2]`, and `r three_lowest[6, 2]`.

Finally, in Africa, the countries with the largest emissions are `r three_highest[13, 2]`, `r three_highest[14, 2]`, and `r three_highest[15, 2]`. The lowest ranking ones are `r three_lowest[15, 2]`,  `r three_lowest[14, 2]`, and  `r three_lowest[13, 2]`.

## 4 GDP per capita (50pt)
Let’s look at GDP per capita (GDP_PC ).

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.

``` {r, message=FALSE}
gdp <- gapminder %>% 
    filter(!is.na(name), !is.na(region), !is.na(GDP_PC), !is.na(time), 
           !is.na(lifeExpectancy), !is.na(totalPopulation)) %>% 
    filter(time == '1960') %>% 
    group_by(name, region, lifeExpectancy, GDP_PC, totalPopulation) %>% 
    select(name, region, time, lifeExpectancy, GDP_PC, totalPopulation)
  
  ggplot(gdp,
         aes(GDP_PC, lifeExpectancy, col=region)) +
    geom_point(aes(size=totalPopulation)) +
    labs( x = 'GDP per capita',
          y = 'Life Expectancy') +
    ggtitle('GDP per capita versus life expectancy by country in 1960')

``` 

I see that the continents with high GDP per capita also have high life expectancy, 
like Americas and Europe. The continents like Africa and Oceania are at the bottom
for both life expectancy and GDP per capita, and while Asia is mainly in the middle 
for life expectancy, it is on the lower side for GDP per capita compared to Americas
and Europe. This suggests a link between GDP per capita and life expectancy, that is
the higher the GDP the higher a continent life expectancy because there is more financial and economic security.



2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r, message=FALSE}
gdp2 <- gapminder %>% 
    filter(!is.na(name), !is.na(region), !is.na(GDP_PC), !is.na(time), 
           !is.na(lifeExpectancy), !is.na(totalPopulation)) %>% 
    filter(time == '2019') %>% 
    group_by(name, region, lifeExpectancy, GDP_PC, totalPopulation) %>% 
    select(name, region, time, lifeExpectancy, GDP_PC, totalPopulation)
  
  ggplot(gdp2,
         aes(GDP_PC, lifeExpectancy, col=region)) +
    geom_point(aes(size=totalPopulation)) +
    labs( x = 'GDP per capita',
          y = 'Life Expectancy') +
    ggtitle('GDP per capita versus life expectancy by country in 2019')
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

The first thing I notice is that Asia has moved up by a lot in both GDP per capita and especially life expectancy. Oceania has also moved higher in life expectancy over the last 60 years. However, Africa is still the lowest in both categories despite the time gap. I think these differences point to significant developments in Asia and to a lesser extent, Oceania and Europe. However, Africa has been left behind and has not developed as much like the other continents. 

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of different size.
```{r, message=FALSE}
average_le_2019 <- gapminder %>% 
    filter(!is.na(lifeExpectancy),!is.na(region), !is.na(name), !is.na(time)) %>% 
    filter(time == '2019') %>% 
    group_by(region, time) %>% 
    summarize(
      total_le = mean(lifeExpectancy)
    ) 
  
  average_le_1960 <- gapminder %>% 
    filter(!is.na(lifeExpectancy),!is.na(region), !is.na(name), !is.na(time)) %>% 
    filter(time == '1960') %>% 
    group_by(region, time) %>% 
    summarize(
      total = mean(lifeExpectancy)
    ) 

```

These results fit what I said about the figures. Between 1960 and 2019, Asia grew in life extectancy and went from being the second lowest in life expectancy to the third lowest. It also experienced more growth in this period than Europe did despite Europe having the highest life expectancy in both years,

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.


```{r message=FALSE}
 average_growth <- gapminder %>% 
    filter(!is.na(lifeExpectancy),!is.na(region), !is.na(name), !is.na(time)) %>% 
    filter(time == '1960'| time == '2019') %>% 
    group_by(region, time) %>% 
    summarize(
      le_average = mean(lifeExpectancy)
    )  %>% 
    mutate(previous_year = lag(le_average), growth = le_average - previous_year) %>% 
    filter(!is.na(growth)) %>% 
    arrange(-growth) %>% 
    knitr::kable() 

average_growth
```

I see that as expected based off previous graphs, Asia has the most growth out of the continents. Surprisingly, Africa has the second-most growth. I believe that's because although it is still at the lowest life expectancy, it did grow a lot from where it was in 1960 to 2019. It just didn't grow enough to surpass other continents' average life expectancy.


6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!

```{r, message=FALSE}
history <- gapminder %>% 
    filter(!is.na(time), !is.na(GDP_PC)) %>% 
    filter(time == '1960'| time == '2019') %>% 
    group_by(time, GDP_PC) %>% 
    summarize(
      GDP_capita = sum(GDP_PC)
    ) 
  
  ggplot(history, aes(x = GDP_PC)) +
    geom_histogram(color = 'purple', alpha=0.5, position = 'identity') +
    facet_wrap(~time)
```


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)
  Hint: check out the function rank()!
  Hint2: 17 for 1960.
```{r, message=FALSE}
rank_1960 <- gapminder %>% 
    filter(!is.na(time), !is.na(name), !is.na(lifeExpectancy)) %>% 
    filter(time == '1960') %>% 
    group_by(time, name, lifeExpectancy) %>% 
    arrange(-lifeExpectancy)
  
  rank_2019 <- gapminder %>% 
    filter(!is.na(time), !is.na(name), !is.na(lifeExpectancy)) %>% 
    filter(time == '2019') %>% 
    group_by(time, name, lifeExpectancy) %>% 
    arrange(-lifeExpectancy)

```
  
The US ranked 17th in terms of life expectancy in 1960, and 46th in 2019.
  
  
8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? Hint: 0.0904 for 1960.

```{r, message=FALSE}
total_countries <- length(rank_1960$name)
  
  filler <- rank_1960 %>% 
    mutate(ranking = rank(desc(lifeExpectancy)),
           ranked = ranking/total_countries) %>% 
    select(name, ranking, ranked) %>% 
    filter(name == 'United States of America')
  
  filler
  
  filler2 <- rank_2019 %>% 
    mutate(ranking = rank(desc(lifeExpectancy)),
           ranked = ranking/total_countries) %>% 
    select(name, ranking, ranked) %>% 
    filter(name == 'United States of America')

  
  filler2
```



9. Finally tell us how many hours did you spend on this PS.

I spent 13 hours.
